%% The first command in your LaTeX source must be the \documentclass command.
%%
%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass{ceurart}

%%
%% One can fix some overfulls
\sloppy

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{listings}
%% auto break lines
\lstset{breaklines=true,mathescape=true,basicstyle={\ttfamily\small},keywordstyle={\ttfamily\bfseries}}
\usepackage{centernot}
\usepackage{graphicx}
\usepackage{proof} % compact proof trees for examples
\usepackage{hyperref} % links
\usepackage{verbatim} % multiline comments
\usepackage[dvipsnames]{xcolor}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{stmaryrd}
%\usepackage{xspace}
\usepackage{color}
\usepackage{amssymb}
\usepackage{pifont}
%\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{pifont}
\usepackage{stmaryrd}
%\usepackage{xspace}
\usepackage{listings}
\usepackage{color}
\usepackage{pgfplots}
\usepackage{subcaption}
\usepackage{cleveref}
%
%%
%% end of the preamble, start of the body of the document source.

% editing macros
\newif\ifsubmit
\submitfalse
%\submittrue

\ifsubmit
 \newcommand{\DA}[1]{{#1}}
 \newcommand{\DAComm}[1]{}
 \newcommand{\AF}[1]{{#1}}
 \newcommand{\AFComm}[1]{}
\else
 \newcommand{\DA}[1]{\textcolor{blue}{#1}}
 \newcommand{\DAComm}[1]{{\scriptsize\textcolor{blue}{[\bf{Angelo: }#1}]}}
 \newcommand{\AF}[1]{\textcolor{red}{#1}}
 \newcommand{\AFComm}[1]{{\scriptsize\textcolor{red}{[\bf{Angelo: }#1}]}}
\fi

\input{macros}

\begin{document}

\conference{ICTCS 2025}

%
\title{On The Space Complexity of Partial Derivatives of Regular Expressions with Shuffle}

\author[1]{Davide Ancona}[orcid=0000-0002-6297-2011,email=davide.ancona@unige.it]

\cormark[1]

\address[1]{University of Genova, Italy}

\author[2]{Angelo Ferrando}[orcid=0000-0002-8711-4670,
 email=angelo.ferrando@unimore.it
]

\address[2]{University of Modena and Reggio Emilia, Italy}


%% Footnotes
\cortext[1]{Corresponding author.}

\begin{keywords}
 partial derivatives \sep
 regular expressions with shuffle \sep
 rewriting-based runtime verification \sep
 space complexity
\end{keywords}

\maketitle
%
\begin{abstract}
 Partial derivatives of regular expressions are an elegant formalism introduced by Antimirov to define an algorithm that generates, from regular expressions, equivalent non-deterministic finite automata (NFA) with a limited number of states.

 While Antimirov's work addresses the classical problem of word recognition in regular languages, here we are interested in runtime verification (RV) of simple properties expressible with regular expressions. In this context, the words to be recognized are finite traces of monitorable events, which define the alphabet of the language, and the corresponding NFA generated from the regular expression may have an intractable number of states.

 This often occurs because the properties to be verified consist of sub-traces of mutually independent events, which are allowed to interleave during the execution of the system under scrutiny (SUS). For this reason, regular expressions used for RV are extended with the shuffle operator. Although regular languages are closed under shuffle, the operator allows for more readable and compact specifications.

 By exploiting partial derivatives, it is possible to follow a rewriting-based approach to RV, where only a single partial derivative needs to be stored at each reduction step, instead of generating a finite automaton with an intractable number of states.

 This raises the question of the space complexity of the largest partial derivative that can be generated. While the upper bound on the total number of generated partial derivatives has been shown to be linear in the size of the initial regular expression, no results can be found in the literature regarding the size of the largest derivative.

 In this paper, we investigate this problem with respect to two different metrics of regular expressions: their height and their total number of nodes when considered as trees. In particular, we show that the height of the largest partial derivative can increase by at most one, while the number of nodes of the largest partial derivative is bounded by $n^2$, where $n$ is the number of nodes of the initial regular expression.

 Surprisingly, these results still hold when regular expressions are extended with shuffle.
\end{abstract}
%
%
\input{regexp}
\input{parDer}
\input{shuffle}
\bibliography{main}
\end{document}
